Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
edits around beta strands

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins.

@@

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence of amino acids.
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
Mathematically, we represent a query sequence as
$\left\{x_{1}...x_{n}\right\}$.
Ideally, if we were writing something like the BioHaskell library, we would
have begun with an algebraic datatype for amino acids deriving
\texttt{Ix}, something like
\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
which would still allow us to index into an array or vector.
However, since we wanted to maintain compatibility with the file format
shared with previous software, we broke with idiomatic Haskell and represented
our amino acids as 
\begin{verbatim}
type AA = Int
\end{verbatim}
and thus our query sequence as
\begin{verbatim}
type querySeq = Vector AA
\end{verbatim}

The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{verbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{verbatim}
to represent the seven allowed state transitions.

Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.
Mathematically, nodes are simply $\left\{a_{s_{k},s_{j}}, e_{s_{j}}(x_{i})\right\}$
for states $s_{k}, s_{j}$ and emission probabilities $e_{s_{j}}(x_{i})$ for a
query sequence element $x_{i}$.
In Haskell, we represent a node as a group of transition probabilities
\begin{verbatim}
data TProbs = TProbs { m_m :: TProb
                     , m_i :: TProb
                     , m_d :: TProb
                     , i_m :: TProb
                     , i_i :: TProb
                     , d_m :: TProb
                     , d_d :: TProb
                     }
\end{verbatim}
and a vector of emission probabilities for the match and insertion states.
\begin{verbatim}
type EProbs = V.Vector logProb
data HmmNode = HmmNode { nodeNum :: Int
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       , transitions :: TProbs
                       }
\end{verbatim}

@@

%% break the viterbi types up and insert into the Plan7 explanation.
%% go over this with NR; I can't make it flow.
% 
% \begin{figure}
% \lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
% \end{figure}
%

@@

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}

@@

A hidden Markov model effectively uses local information about
amino-acid sequences.
But because real proteins are three-dimensional structures,
their amino-acid sequences are subject to \emph{nonlocal}
constraints. 
In~particular, when a real protein folds, groups of amino acids
that are far away in the one-dimensional sequence can nevertheless be
adjacent in three-dimensional space.
Moreover, two such groups can become hydrogen-bonded to one another,
effectively ``stuck together.''
Such groups are called \emph{beta strands}.
\remark{I hope this is relatively accurate. --NR}

@@

 simply implements the Viterbi algorithm for computing the optimal parse 
of a protein sequence onto a hidden Markov model.
Thus, the graphical model of 
a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
In 
contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
SMURFLite~\cite{smurflite}, our previous work.
Like SMURF and SMURFLite, MRFy 
pays attention to longer-range structural interactions than a simple hidden 
Markov model can understand.
In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
Beta strands are regions of protein 
structure that ``stick'' to one another.
Because of their pairwise 
interactions, they are highly conserved over the course of evolution.
Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

@@

As inexperienced functional programmers developing a complex piece of
software in Haskell, we encountered some barriers.
The first of these was profiling with GHC.
We previously mentioned that profiling with GHC required annotating
cost centers by hand, producing ugly code.
Beyond that, we also found it very difficult to profile code at all.
In particular, profiling required that all installed libraries be re-compiled
with profiling support,which was a process that we could not replicate
consistently.

Another barrier we faced was how to test our code using QuickCheck.
Indeed, the barrier was so formidable in our case that we still have not been 
able to use it.
The problems we faced specifically were two-fold: How do we start testing
a complex module? How can we use QuickCheck to generate random inputs with 
complex constraints? 
We wished to test invariants around beta-strand placement 
(potential solutions in our stochastic search), but found the QuickCheck generator
to be difficult to write.

Finally, one last barrier we faced was debugging runtime errors.
Runtime errors in the form of out-of-bounds indexing were particularly 
difficult to track down because they were not accompanied by a stack trace.
(While profiling with GHC could help remedy this, profiling was itself a 
barrier, as previously mentioned.)
We had to resort to polluting our code with calls to the \texttt{trace} function 
in order to track down where the out-of-bounds errors were occurring.

@@

\subsection{What you must know to succeed}
If others would like to follow in our footsteps, there are a few
things we believe one must know in order to succeed.
Firstly, some experience with functional programming concepts is a necessity. 
Specifically, coming from an imperative background, we found transforming input data 
from the HMMER file format without the ability to mutate state was difficult.
In contrast, writing ``pure math'' like the Viterbi algorithm felt easier
and more natural than it would have been in C++.

Secondly, identifying existing tools and libraries from Hackage that are mature 
is also important.
We found that some packages that could have been immensely useful---like one 
that provides backtraces---were either not maintained or contained bugs.
Some other packages like Pads were functional but immature, which resulted in 
additional code to work around the parts of Pads that had not been implemented 
yet. 
In contrast, other packages such as \texttt{Data.Memocombinators} were
useful and mature.

@@

\subsection{Evaluation}
\remark{I think we need more in this section. ---AG}
In hindsight, we feel that there are several key facets of our project in 
Haskell that deserve evaluation.
Firstly, we found that existing tools and libraries were plentiful and 
typically helped us greatly.
In particular, \texttt{Data.Memocombinators} made dynamic 
programming with Viterbi exquisitely simple and elegant.
\texttt{Control.Parallel.Strategies} also made the parallelization of Viterbi trivial.
Most of the tools we used came straight from Hackage, and although 
documentation of some form was almost always present, it was sometimes 
difficult to tell if a particular tool was right for us.
In particular, parts of the documentation were inaccessible to us as beginning 
functional programmers.
Documentation that was accessible to us typically contained straight-forward 
examples showing how to use that particular tool or library.

>>>>

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins. 
\remark{``Homology'' means ``share a common ancestor.  I hope we can
avoid this story. ---NR}

@@

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
\remark{Time to introduce the elements of Viterbi's equations!}
The match state determines the probability that the sequence under
test has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model.
Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.

We employed a variety of Haskell data types to represent the hidden Markov
model.
Ideally, if we were writing something like the BioHaskell library, we would
have begun with an algebraic datatype for amino acids deriving
\texttt{Ix}, something like
\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
which would still allow us to index into an array or vector.
However, since we wanted to maintain compatibility with the file format
shared with previous software, we broke with idiomatic Haskell and represented
our amino acids as \texttt{Int}.
Similarly, to avoid a great deal of data transformation, each state's emission
and insertion probabilities were simply
\begin{verbatim}type EProbs = V.Vector Double \end{verbatim} while the states themselves
were algebraic datatypes: 
\begin{verbatim}data HMMState = Match | Insert | Delete \end{verbatim}

\emph{[HOW ABOUT SOME HASKELL TYPE DEFINITIONS AT OR NEAR THIS POINT?
    I THINK WE'RE READY.]}

% time for types.
% first, ideal: amino acids as an algebraic DT
% next, back off this: we're working with an existing input format
% so, how is the graph represented (vector of state-transitions, 
% vector of emissions)
% so show those types

@@

%% break the viterbi types up and insert into the Plan7 explanation.
%% go over this with NR; I can't make it flow.

\begin{figure}
\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
\end{figure}

@@

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}I_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}I_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}

@@

A hidden Markov model effectively uses local information about
amino-acid sequences.
But real proteins are three-dimensional structures,
and their amino-acid sequences are subject to \emph{nonlocal}
constraints. 
And when a real protein folds, groups of amino acids
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
We~are studying adjacent groups called \emph{beta strands}, which
can become hydrogen-bonded to each other,
and are effectively ``stuck together.''

@@

In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
Because of their pairwise 
interactions, they are highly conserved over the course of evolution.
Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

@@

As inexperienced functional programmers, developing a complex piece of
software in Haskell posed several barriers.
\remark{NOAH: You did the profiling, could you make sure what I've said here
is accurate and/or relevant? I didn't know what else to say. ---AG}
The first of which was profiling with GHC.
We previously mentioned that profiling with GHC produced ugly code with
cost center annotations scattered throughout our code, but we also found it
very difficult to profile code at all.
In particular, profiling required that all installed libraries be re-compiled
with profiling support---which was a process that we could not replicate
consistently.

Another barrier we faced was how to test our code using QuickCheck.
Indeed, the barrier was so formidable in our case that we still have not been 
able to use it.
The problems we faced specifically were two-fold: How do we start testing
a complex module? and How can we use QuickCheck to generate random inputs with 
complex constraints?

Finally, one last barrier we faced was debugging runtime errors.
Runtime errors in the form of out-of-bounds indexing were particularly 
difficult to track down because they were not accompanied by a stack trace.
(While profiling with GHC could help remedy this, profiling was itself a 
barrier, as previous mentioned.)
We had to resort to polluting our code with cals to the \texttt{trace} function 
in order to track down where the out-of-bounds errors were occurring.

@@

\subsection{What you must know to succeed}
\remark{I don't like this. I really struggled with this section. ---AG}
If others would like to follow in our footsteps, there are a few
things we believe one must know in order to succeed.
Firstly, some experience with functional programming concepts is a necessity. 
Specifically, coming from an imperative background, writing algorithms and 
modeling data without the ability to mutate state were the most difficult 
concepts to adapt to.

Secondly, identifying existing tools and libraries from Hackage that are mature 
is also important.
We found that some packages that could have been immensely useful---like one 
that provides backtraces---were either not maintained or contained bugs.
Some other packages like Pads were functional but immature, which resulted in 
additional code to work around the parts of Pads that had not been implemented 
yet.

@@

\subsection{Evaluation}
\remark{I think we need more in this section. ---AG}
In hindsight, we feel that there are several key facets of our project in 
Haskell that deserve evaluation.
Firstly, we found that existing tools and libraries were plentiful and 
typically helped us greatly.
In particular, Data.Memocombinators made dynamic 
programming with Viterbi exquisitely simple and elegant.
Control.Parallel.Strategies also made the parallelization of Viterbi trivial.
Most of the tools we used came straight from Hackage, and although 
documentation of some form was almost always present, it was sometimes 
difficult to tell if a particular tool was right for us.
In particular, parts of the documentation were inaccessible to us as beginning 
functional programmers.
Documentation that was accessible to us typically contained straight-forward 
examples showing how to use that particular tool or library.
